### 2009-05-19 paper.tex
Finalizing edits...

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies, where we kept the branch lengths of the two
sibling species constant and varied the branch length of the
outgroup.  We fixed the tree topology as $((X:1,Y:1):1,Z:z)$, where
the branch length $z$ of the outgroup was varied between 0.5 and 3.0
and a branch length of 1 corresponds to one expected substitution per site
in loop sequence.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$, loops of length $\in [3, 10]$ and
stems of lengths $\in [1, 20]$.

We expected to find a ``transition point'' in our reconstruction
accuracy near the point $z \simeq 1$, corresponding to an outgroup
distance equal to the distance from either of the sibling species to
the most recent common ancestor.  Our results, shown in
\figref{simulations}, indicate such a transition point for alignment
accuracy but not for ancestral reconstruction accuracy.
% to do: here now

\cite{Mossel2003}

While these results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model, recent results
demonstrating that inference of sequence alignment is surprisingly
robust to mis-parametrized models \cite{Lunter2007} suggest that the same
may be true for inference of ancestral structures.

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of what are currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored.
One promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

@@

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/simulations.pdf}
  \caption{
    \textbf{Dependence of reconstruction accuracy on outgroup branch
      length.}
    We allowed the outgroup branch length $z$ to vary within $[0.5,
    3.0]$, where a branch length of 1 corresponds to one expected
    substitution per site in loop sequence.  We divided this interval
    into 25 bins and simulated 25 alignments for each bin.
    Parameters used were:
  }
  \figlabel{simulations}
\end{figure}

>>>>

% figures in figs_plos:
% 1: parent.pdf
% 2: threeway.pdf
% 3: stree-with-structure.pdf
% 4: stree-mutations-with-structures.pdf
% 5: simulations.pdf

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies, where we kept the branch lengths of the two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$, loops of length $\in [3, 10]$ and
stems of lengths $\in [1, 20]$.

The results of our simulations, illustrated in \figref{simulations},
demonstrate that the accuracy of both structural
reconstruction and alignment accuracy depend on the outgroup
branch length.  However, while the difficulty of alignment inference
increases in proportion to the outgroup branch length, becoming
near-impossible for very long outgroup branch lengths, the same need not
true for structural reconstruction accuracy.  As the outgroup branch
length increases, its structure becomes increasingly decoupled from
that of its ancestor, allowing the algorithm to ``ignore'' the
information which it contains.
Similarly, while we can perfectly reconstruct the ancestral
structure given an outgroup branch length of 0 (the trivial case), we
cannot perfectly reconstruct the alignment.

While these results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model, recent results
demonstrating that inference of sequence alignment is surprisingly
robust to mis-parametrized probabilistic models \cite{Lunter2007}
suggest that the same may be true for probabilistic inference of
ancestral structures.

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of what are currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored, such as the
progressive reconstruction used by the \ortheus\ program for
reconstructing ancestral genomes \cite{PatenEtAl2008}.
Another promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

@@

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/simulations.pdf}
  \caption{
    \textbf{Dependence of reconstruction accuracy on outgroup branch
      length.}
    We investigated the dependence of structural reconstruction and
    alignment accuracy on outgroup branch length by simulating the
    evolution of three structural RNAs under our model.  The two
    sister species were equidistant from the ancestral sequence, with
    a branch length of 1, corresponding to one expected
    substitution per site in loop sequence.  We allowed the outgroup
    branch length to vary between $[0, 2.5]$.  We divided this interval
    into 25 bins and simulated 25 alignments for each bin.
    Parameters used were $\lambda_1 = 0.025$ and $\mu_1 = 0.03$ for
    loop sequence and $\lambda_2 = 0.007$ and $\mu_2 = 0.01$ for stem
    sequence; the probability of a stem insertion was $0.1$.  These
    yielded a mean loop length of 5 and a mean stem length of 2.33,
    with $\sim 0.3$ substructure indels per alignment.
    We selected challenging alignments to reconstruct by requiring
    that there be at least two ancestral stems, loops of length $\in [3,
    10]$ and stems of length $\in [1, 20]$; to reduce the complexity of
    our algorithms we additionally required that the sequences have
    lengths $\in [30, 70]$.
 }
  \figlabel{simulations}
\end{figure}
